---
title: "Final Project"
author: "Anish Gupta"
date: "2023-11-07"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(plm)
library(lmtest)
library(ggplot2)
library(tidyr)
library(applicable)
library(corrplot)
library(lubridate)
library(stargazer)
library(Metrics)
library(car)
library(sjPlot)
library(sjmisc)
library(sjlabelled)
options(scipen=999) 
knitr::opts_chunk$set(echo = TRUE)
```


#### Read College Completion Datasets
```{r}
setwd("C:\\Users\\anish\\Documents\\Applied Multivariate Methods\\databeats-college-completion\\databeats-college-completion\\data")

institution_details <- read_csv("cc_institution_details.csv")

institution_grads <- read_csv("cc_institution_grads.csv")

```

#### Data Wrangling for institution_details
```{r}
#From the institution details dataframe, filter and keep 4-year institutions.
four_year_institution_details <- institution_details %>%
  filter(level == "4-year")

#Check to see if the control binary variable is Public, then assign a value of 1, else assign a value of 2.
four_year_institution_details$control_public_private = ifelse(four_year_institution_details$control == "Public", 1, 2)

dim(four_year_institution_details)
four_year_institution_details %>%
  summarise(sum_grad_100_value = sum(is.na(grad_100_value)),
            sum_grad_150_value = sum(is.na(grad_150_value)),
            sum_med_sat_value = sum(is.na(med_sat_value)),
            sum_aid_value = sum(is.na(aid_value)),
            sum_pell_percentile = sum(is.na(pell_percentile)),
            sum_retain_percentile = sum(is.na(retain_percentile)),
            sum_ft_pct = sum(is.na(ft_pct)),
            sum_endow_value = sum(is.na(endow_value)),
            sum_fte_percentile = sum(is.na(fte_percentile)),
            sum_awards_per_state_value = sum(is.na(awards_per_state_value)),
            sum_awards_per_natl_value = sum(is.na(awards_per_natl_value)),
            sum_control_public_private = sum(is.na(control_public_private)))
```

#### Create a subset of the necessary variables and drop med_sat_value, endow_value.
```{r}
four_year_institution_details_subset <- four_year_institution_details %>%
  select(unitid, grad_100_value, grad_150_value, control_public_private, aid_value, pell_percentile, retain_percentile, ft_pct, fte_percentile, awards_per_state_value,awards_per_natl_value)

dim(four_year_institution_details_subset)
```

#### Drop remaining rows with NA values
```{r}
#For this project, drop the NA values.
clean_four_year_institution_details <- four_year_institution_details_subset %>% drop_na()

dim(clean_four_year_institution_details)
```
#### Wrangle institution_grads
```{r}
#table(institution_grads$cohort)

institution_grads %>%
  summarise(sum_institution_grads_year = sum(is.na(year)),
            sum_institution_grads_gender = sum(is.na(gender)),
            sum_institution_grads_race = sum(is.na(race))
            )

#Filter for Bachelor's/equivalent-seeking cohort at 4-year institutions
four_year_institution_grads <- institution_grads %>%
  filter(cohort == "4y bach",
         gender != "B",
         race != "X")
four_year_institution_grads <- four_year_institution_grads %>% 
  group_by(unitid) %>% 
  filter(year == max(year))

#Create a subset of the necessary variables.
four_year_institution_grads_subset <- four_year_institution_grads %>%
  select(unitid, gender, race, grad_cohort)

#For this project, drop the NA values.
clean_four_year_institution_grads <- four_year_institution_grads_subset %>% drop_na()

clean_four_year_institution_grads_percent <- clean_four_year_institution_grads %>%
  group_by(unitid) %>%
  mutate(grad_cohort_percent = (grad_cohort / sum(grad_cohort))*100)

```



```{r}
institution_grad_details <- clean_four_year_institution_details %>% left_join( clean_four_year_institution_grads_percent, 
        by=c('unitid'='unitid'))

summary(institution_grad_details)
institution_grad_details <- institution_grad_details %>% drop_na()
summary(institution_grad_details)
stargazer(as.data.frame(institution_grad_details),header=FALSE, type = 'text')

```


#### Pivot the merged dataframe wider
```{r}
institution_grad_details_wider <- institution_grad_details %>% 
 select(-grad_cohort) %>%
  pivot_wider(
    names_from = c(gender, race), 
    values_from = grad_cohort_percent
  )

```

#### Examine the data and run summary statistics
```{r}
summary(institution_grad_details_wider)
stargazer(as.data.frame(institution_grad_details_wider), median = TRUE, type = 'text')
sum(is.na(institution_grad_details_wider))
```
```{r}
plot(institution_grad_details_wider$grad_100_value,institution_grad_details_wider$aid_value )
plot(institution_grad_details_wider$grad_150_value,institution_grad_details_wider$aid_value )
plot(institution_grad_details_wider$grad_100_value,institution_grad_details_wider$retain_percentile)
plot(institution_grad_details_wider$grad_150_value,institution_grad_details_wider$retain_percentile)
plot(institution_grad_details_wider$grad_100_value,institution_grad_details_wider$ft_pct)
plot(institution_grad_details_wider$grad_150_value,institution_grad_details_wider$ft_pct)
plot(institution_grad_details_wider$grad_100_value,institution_grad_details_wider$fte_percentile)
plot(institution_grad_details_wider$grad_150_value,institution_grad_details_wider$fte_percentile)
plot(institution_grad_details_wider$grad_100_value,institution_grad_details_wider$awards_per_state_value)
plot(institution_grad_details_wider$grad_150_value,institution_grad_details_wider$awards_per_state_value)
plot(institution_grad_details_wider$grad_100_value,institution_grad_details_wider$awards_per_natl_value)
plot(institution_grad_details_wider$grad_150_value,institution_grad_details_wider$awards_per_natl_value)

```

#### Create Histogram for grad_100_value and grad_150_value and all other variables
```{r}
hist(institution_grad_details_wider$grad_100_value)
hist(institution_grad_details_wider$grad_150_value)

ggplot(gather(institution_grad_details_wider[,-1]), aes(value)) +
  geom_histogram(bins = 10) +
  facet_wrap(~key, scales = "free_x")
```
```{r}
hist(log(institution_grad_details_wider$grad_100_value))
hist(institution_grad_details_wider$aid_value, breaks = 100)
hist(log(institution_grad_details_wider$aid_value), breaks = 100)
hist(institution_grad_details_wider$F_A, breaks = 100)
hist(institution_grad_details_wider$F_Ai, breaks = 100)
hist(institution_grad_details_wider$F_B, breaks = 100)
hist(institution_grad_details_wider$F_H, breaks = 100)
hist(institution_grad_details_wider$M_A, breaks = 100)
hist(institution_grad_details_wider$M_Ai, breaks = 100)
hist(institution_grad_details_wider$M_B, breaks = 100)
hist(institution_grad_details_wider$M_H, breaks = 100)
```

The distribution for the four year graduation is slightly right-skewed. It also has an outlier where a small number of schools the four year graduation rate is more than 80%. A log transformation here might be helpful.
Upon log transformation we see that the variable becomes left skewed and therefore it will not be useful.

The distribution for the six year graduation is normal. 


The F_A, F_Ai, F_B, M_A, M_Ai, M_B, F_H, M_H are zero inflated distributions and a log transformation will not help here.

aid_value is not zero inflated distribution and therefore may benefit from log transformation.
Upon log transformation we see that the variable is now more normally distributed than before.

aid_value will be log transformed in the analysis


#### Boxplot for grad_100_value and grad_150_value
```{r}
boxplot(institution_grad_details_wider$grad_100_value)
boxplot(institution_grad_details_wider$grad_150_value)
```

#### The boxplot for the four year graduation rate has a couple of outliers.

#### The boxplot for the six year graduation rate is normally distributed and contains no outliers.

#### Correlation Matrix
```{r}
#glimpse(institution_grad_details_wider)
cor_matrix <- cor(institution_grad_details_wider[,c(-1,-10,-11)])
cor_matrix

corrplot(cor(institution_grad_details_wider[,c(-1,-10,-11)]))
```


#### Run the first multiple linear regression model for 4-year graduation with all variables
```{r}
# Log transform the variables
institution_grad_details_wider$ln_aid_value <-log(institution_grad_details_wider$aid_value)

model_1 <- lm(grad_100_value~control_public_private+ln_aid_value+pell_percentile+retain_percentile+ft_pct+fte_percentile+M_W+F_W+M_B+F_B+M_H+F_H+M_Ai+F_Ai+M_A, data = institution_grad_details_wider)

alias(model_1)
summary(model_1)
stargazer(model_1, title="Results", align=TRUE,type = "text")
```

#### Checking the regression assumptions for Model 1 and create histogram
```{r}
institution_grad_details_wider$residuals_1 <- residuals(model_1)
hist(institution_grad_details_wider$residuals_1)
institution_grad_details_wider$predvals_1<-fitted(model_1)
plot(institution_grad_details_wider$predvals_1,institution_grad_details_wider$residuals_1)
plot(institution_grad_details_wider$pell_percentile,institution_grad_details_wider$residuals_1)
plot(institution_grad_details_wider$control_public_private,institution_grad_details_wider$residuals_1)
plot(institution_grad_details_wider$ln_aid_value,institution_grad_details_wider$residuals_1)
plot(institution_grad_details_wider$retain_percentile,institution_grad_details_wider$residuals_1)
plot(institution_grad_details_wider$ft_pct,institution_grad_details_wider$residuals_1)
plot(institution_grad_details_wider$fte_percentile,institution_grad_details_wider$residuals_1)
plot(institution_grad_details_wider$M_W,institution_grad_details_wider$residuals_1)
plot(institution_grad_details_wider$F_W,institution_grad_details_wider$residuals_1)
plot(institution_grad_details_wider$M_B,institution_grad_details_wider$residuals_1)
plot(institution_grad_details_wider$F_B,institution_grad_details_wider$residuals_1)
plot(institution_grad_details_wider$M_Ai,institution_grad_details_wider$residuals_1)
plot(institution_grad_details_wider$F_Ai,institution_grad_details_wider$residuals_1)
plot(institution_grad_details_wider$M_A,institution_grad_details_wider$residuals_1)
plot(institution_grad_details_wider$M_H,institution_grad_details_wider$residuals_1)
plot(institution_grad_details_wider$F_H,institution_grad_details_wider$residuals_1)

vif(model_1)
stargazer(vif(model_1), title="VIF", type = "text")
```
#### Checking to see if a quadratic regression model will improve model fit and will fix heteroskedasticity
```{r}
model_1.1 <- lm(grad_100_value~control_public_private+ln_aid_value+pell_percentile+pell_percentile^2+retain_percentile+retain_percentile^2+ft_pct+ft_pct^2+fte_percentile+fte_percentile^2+M_W+F_W+M_B+F_B+M_H+F_H+M_Ai+F_Ai+M_A, data = institution_grad_details_wider)

summary(model_1.1)
stargazer(model_1.1, title="Results", align=TRUE,type = "text")
```

#### Checking the regression assumptions for Model 1.1 and create histogram
```{r}
institution_grad_details_wider$residuals_1.1 <- residuals(model_1.1)
hist(institution_grad_details_wider$residuals_1.1)
institution_grad_details_wider$predvals_1.1<-fitted(model_1.1)
plot(institution_grad_details_wider$predvals_1.1,institution_grad_details_wider$residuals_1.1)
vif(model_1.1)
stargazer(vif(model_1.1), title="VIF", type = "text")
```



#### Run the second multiple linear regression model by dropping ln_aid_value, awards_per_natl_value, F_B, M_B, M_Ai, F_Ai, M_A,F_H due to multicolinearity
```{r}
model_2 <- lm(grad_100_value~control_public_private+pell_percentile+retain_percentile+ft_pct+fte_percentile+F_W +F_A, data = institution_grad_details_wider)
summary(model_2)
stargazer(model_2, title="Results", align=TRUE,type = "text")
```

#### Checking the regression assumptions for Model 2 and create histogram
```{r}
institution_grad_details_wider$residuals_2 <- residuals(model_2)
hist(institution_grad_details_wider$residuals_2)
institution_grad_details_wider$predvals_2<-fitted(model_2)
plot(institution_grad_details_wider$predvals_2,institution_grad_details_wider$residuals_2)
vif(model_2)
stargazer(vif(model_2), title="VIF", type = "text")
```
#### Multiple Linear Regression for 6-year graduation with all variables
```{r}
model_3 <- lm(grad_150_value~control_public_private+ln_aid_value+pell_percentile+retain_percentile+ft_pct+fte_percentile+M_W+F_W+M_B+F_B+M_H+F_H+M_Ai+F_Ai+M_A, data = institution_grad_details_wider)
summary(model_3)
stargazer(model_3, title="Results",type = "text")
```
#### Hist for model 3
```{r}
institution_grad_details_wider$residuals_3 <- residuals(model_3)
hist(institution_grad_details_wider$residuals_3)
institution_grad_details_wider$predvals_3<-fitted(model_3)
plot(institution_grad_details_wider$predvals_3,institution_grad_details_wider$residuals_3)
vif(model_3)
stargazer(vif(model_3), title="VIF", type = "text")
```

#### Run the fourth multiple linear regression model for 6 year graduation by dropping ln_aid_value, F_B, and M_B due to multicolinearity
```{r}
model_4 <- lm(grad_150_value~control_public_private+retain_percentile+ft_pct+fte_percentile+M_W+F_W+M_H+F_H+M_Ai+F_Ai+M_A+F_A, data = institution_grad_details_wider)
summary(model_4)
stargazer(model_4, title="Results",type = "text")
```

#### Histogram for model 4
```{r}
institution_grad_details_wider$residuals_4 <- residuals(model_4)
hist(institution_grad_details_wider$residuals_4)
institution_grad_details_wider$predvals_4<-fitted(model_4)
plot(institution_grad_details_wider$predvals_4,institution_grad_details_wider$residuals_4)
vif(model_4)
stargazer(vif(model_4), title="VIF", type = "text")
```

#### PCA
```{r}
#glimpse(institution_grad_details_wider)
my_pca <- prcomp(institution_grad_details_wider[,c(-1,-2,-3,-5,-10,-11,-23,-24,-25,-26,-27,-28,-29,-30,-31,-32)], scale = TRUE)
summary(my_pca)
```
Here, only PC1, PC2, PC3, PC4, PC5, PC6, PC7 have an eigenvalue > 1.


#### Loading Scores
```{r}
loading_scores <- my_pca$rotation
loading_scores
```

#### Scree Plot
```{r}
var <- my_pca$sdev^2

#in base R
plot(var, type = "b", lty = 1)

#using ggplot
qplot(c(1:16), var) +
  geom_line() +
  geom_point(aes(size=1))+
  xlab("Principal Component") +
  ylab("Eigenvalues") +
  ggtitle("Scree Plot") #+
  #ylim(0, 2.5)
```

#### Scree Plot by bar graph
```{r}
#CREATE A SCREE-PLOT OF SHARE OF VARIANCE 
var_pct <- var/sum(var)*100

#barplot in base R
barplot(var_pct, main="Scree Plot - Share of Var.",
        xlab="Principal Components",
        ylab="Share of Variance Explained")
```

#### Biplot
```{r}
#CREATE BIPLOT FOR PC1 AND PC2
biplot(my_pca, cex=.4)
```

#### Present a table of the scores for the first six observations in your original dataset.
```{r}
scores <- my_pca$x
head(scores, n=6)
```

#### Using these new PCA scores, regress the principal components you elected to retain on your dependent variable from your original dataset.
```{r}
scores_combined <- cbind(institution_grad_details_wider,scores)

reg_1 <- lm(grad_100_value ~ PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7, data=scores_combined)
summary(reg_1)
stargazer(reg_1, title="Results",type = "text")

reg_2 <- lm(grad_150_value ~ PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7, data=scores_combined)
summary(reg_2)
stargazer(reg_2, title="Results",type = "text")
```
```{r}
scores_combined$residuals_reg_1 <- residuals(reg_1)
hist(scores_combined$residuals_reg_1)
scores_combined$predvals_reg_1<-fitted(reg_1)
plot(scores_combined$predvals_reg_1,scores_combined$residuals_reg_1)
```
```{r}
scores_combined$residuals_reg_2 <- residuals(reg_2)
hist(scores_combined$residuals_reg_2)
scores_combined$predvals_reg_2<-fitted(reg_2)
plot(scores_combined$predvals_reg_2,scores_combined$residuals_reg_2)
```
```{r}
#glimpse(scores_combined)
scores_combined %>%
  filter(control_public_private == 2) %>%
  summarise(median_grad_100_private = median(grad_100_value),
            median_grad_150_private = median(grad_150_value))

scores_combined %>%
  filter(control_public_private == 1) %>%
  summarise(median_grad_100_public = median(grad_100_value),
            median_grad_150_public = median(grad_150_value))
```

